\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-02-01 doc/publications/plos16/plos16.tex
some small changes to results section

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables (features, p) of which 5 
are the response variable and 5000 samples (n), as discussed in the methods section.

We fit the random forest model to obtain the classification accuracy by capturing the out-of-bag error (OOB) and perform the variable importance analysis where we expect to retrieve the
5 features in order of the weight that was assigned to them, which is captured by the rank-biased-overlap (RBO) \cite{Webber2010} measure.
We are running Apache Spark 1.6.1 on a YARN cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.

@@

As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables selected at each node) does not result in a 
good classification performance for this large feature dataset. Note, that the plot shows the proportion of $n/p$ for \mtry\ with the default being $\sqrt{p} \approx 1581$. 
The OOB  for this value of \mtry does not exceed 0.5 on a scale from 0 (perfect) to 1 (random), even when the number of trees are increased (\ntree).

@@

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
      \caption{OOB classification error.} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
      \caption{The rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
      the better result, that is, a lower OOB or a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

@@

\marginpar{tables 2 has been commented out -- there is no data in it. Piotr could add the data or we could remove both
  tables just leave the figures. I have done this for the time being} 
Next we investigate \cursedforest scalability with
regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
axes and the values on y axes are expressed as trees per hour).

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest can evaluate variable importance for millions of features}
\label{section:synthetic}
In this section we demonstrate the two features of \cursedforest, namely: variable selection analysis to select the most
associated set of features amongst millions of dimensions; and its ability to accurately classify the labels of
high-dimensional samples.  We therefore generate a synthetic dataset with 2.5 million variables ($p=2500000$) of which 5 
are related to the response variable and 5000 samples ($n=5000$), as discussed in the methods section.

We fit the random forest model and estimate the classification accuracy by capturing the out-of-bag error (OOB) and the
variable selection performance by capturing the rank-biased-overlap (RBO) \cite{Webber2010} measure. We expect to
retrieve the 5 features in order of the weight that was assigned to them and with a higher variable importance measure
than the noise variables. RBO provides a quantiative measure of this.  We are running Apache Spark 1.6.1 on a YARN
cluster with 12 worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.

@@

As shown in Fig~\ref{figure:out-of-bag-prediction-error-prod.png} the default value for \mtry (number of variables
selected at each node) does not result in a good classification performance for this large feature dataset. Note, that
the plot shows the proportion of $n/p$ for \mtry\ with the default being $\sqrt{p} \approx 1581$ ($n/p=0.0006$).  The
OOB for this value of \mtry does not drop below 0.5, even when the number of trees are increased (\ntree).

@@

\begin{figure}[tbhp] 
  \begin{adjustwidth}{-1.00in}{0in}
    \caption{\textbf{Out of bag classification error and rank-biased-overlap estimates}} 
    \label{figure:synth1}
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/oob_levleplot.png}
      \caption{OOB classification error.} 
      \label{figure:out-of-bag-prediction-error-prod.png} 
%      \vspace{4ex}
    \end{subfigure} 
    \begin{subfigure}[b]{0.5\linewidth}
      \centering
      \includegraphics[totalheight=8cm]{./figs/rbo_levleplot.png}
      \caption{The rank-biased-overlap measure.} 
      \label{figure:rbo-prod.png} 
%      \vspace{4ex}
    \end{subfigure} 
    \begin{flushleft} 
      Random forests with 100 trees were trained on the synthetic dataset of 2.5M features and 5K samples.
      Each measure is shown as a function of \mtry and \ntree. Note that the plot colors have been chosen so that the lighter color indicates
      the better result, that is, a lower OOB or a higher RBO.
    \end{flushleft}
  \end{adjustwidth}
\end{figure}

@@

Next we investigate \cursedforest scalability with
regards to the size of data, by varying the number of variables and sample for a fixed \mtry fraction of 0.25 and
execution of 128 CPU cores. The results are visualized in Fig~\ref{figure:synth2} below (please note log scale on the
axes and the values on y axes are expressed as trees per hour).
